---
title: 'Expression backend benchmark: embedded vs bpcells vs h5'
author: 'Xuesong Wang'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Expression backend benchmark: embedded vs bpcells vs h5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

`exportFromSeurat()` and `convertSeuratToCerebro()` can persist the expression matrix in one of three modes via `expression_matrix_mode`: `"embedded"` (default — matrix lives inside the `.crb`), `"bpcells"` (BPCells on-disk matrix in a sibling `<stem>.bpcells/` directory), or `"h5"` (sparse CSC HDF5 in a sibling `<stem>.h5` file, written via `HDF5Array::writeTENxMatrix()` since 1.7.0). The same crb can therefore expand into very different on-disk footprints, startup costs, and runtime memory profiles, and the right choice depends on what you optimise for.

This vignette walks through a controlled comparison of all three backends on **synthetic data** generated by `tests/smoke/src/00_prepare_synthetic_data.R`. Using synthetic data means every number below is fully reproducible — no external datasets required.

# The three backends

All three store the same matrix; what differs is **where** it lives, **when** it gets read, and what's left on disk and in RAM. The right pick depends on what you optimise for.

**Eager vs lazy in one paragraph.** _Eager_ = "read the whole matrix into RAM up front, then index it from there" — the `embedded` and (1.6.0) h5 paths. _Lazy_ = "keep the matrix on disk, only read the slice the caller asked for" — the `bpcells` and (1.7.0) h5 paths. Cerebro's typical query is "give me one gene across all cells", which touches < 0.003 % of the matrix — so paying gigabytes of RAM and seconds of attach time to read everything eagerly is mostly wasted work. Lazy backends keep the OS page cache hot for recently-read columns, so repeat queries are essentially memory-speed without ever committing the whole matrix to the R process's RAM. The `crb$expression <- NULL` step in the exporter is what makes this work: leave the in-memory `dgCMatrix` attached and `saveRDS` would re-embed it inside the `.crb`, defeating the whole point of the sibling.

| dimension | `embedded` | `bpcells` | `h5` |
|---|---|---|---|
| **Where the matrix lives** | inside the `.crb` itself, serialised with `saveRDS` | sibling `<stem>.bpcells/` directory (BPCells's own packed format) | sibling `<stem>.h5` file (10X TENx CSC layout, written by `HDF5Array::writeTENxMatrix()`) |
| **What the `.crb` contains** | metadata + projections + trees + **the full matrix** | metadata + projections + trees + a lightweight `IterableMatrix` handle | metadata + projections + trees + a backend tag (`type = "h5"`, `location = "<stem>.h5"`) |
| **Attach behaviour** | nothing to do; matrix is already in memory after `readRDS` | `open_matrix_dir()` opens file handles on the sibling dir — no data read yet | `TENxMatrix(file, group)` reads only top-level dimnames; the `data`/`indices`/`indptr` arrays stay on disk |
| **Runtime data access** | direct `dgCMatrix` indexing (in RAM) | streams chunks from disk through BPCells's bit-packed integer format (when integer counts; raw double otherwise) | streams individual TENx columns from disk through `HDF5Array`'s page cache (column = single gene; OS caches recently-read columns automatically) |
| **Best for** | small datasets, single-file portability, RAM is plentiful | RAM-constrained host with very large matrices, batched/chunked workloads | most deployments — fastest startup, lowest RAM, fastest per-gene queries |
| **Extra package needed** | none | `BPCells` (GitHub) | `HDF5Array` (Bioconductor; pulls `rhdf5` transitively) |
| **Portability** | single `.crb` file; works with any reader | `.crb` + `<stem>.bpcells/` must travel together | `.crb` + `<stem>.h5` must travel together; the `.h5` is readable by any TENx-compatible tool |

The two external backends (`bpcells`, `h5`) are both lazy — the in-memory `dgCMatrix` is never materialised on attach. The 1.7.0 lazy h5 attach reuses Roman Hillje's original `HDF5Array::TENxMatrix` seed pattern from [`create_expression_matrix_in_h5_format.Rmd`](create_expression_matrix_in_h5_format.html); prior to 1.7.0 the h5 attach was eager and recreated a full `dgCMatrix` from raw rhdf5 reads at attach time (see the [Appendix](#appendix-1-6-0-eager-h5-baseline) for the before/after numbers).

**Disk-size caveat — it depends on fixture size and value type.** HDF5 default-gzips the TENx CSC arrays (`writeTENxMatrix` ships with `level = 4`). BPCells's on-disk format only bit-packs when the matrix's storage type is integer — `dgCMatrix@x` is `double` even when every value is an integer count, so prior to 1.7.0 the bpcells exporter wrote raw doubles. Since 1.7.0 the bpcells exporter auto-detects losslessly-integer values and calls `BPCells::convert_matrix_type("uint32_t")` before write, putting bpcells sibling size on par with h5 (~5× smaller than the raw-double baseline). On _small or dense_ fixtures the h5 vs embedded picture flips: HDF5 chunk + dimnames overhead can exceed the compressed payload, so the h5 sibling ends up larger than the embedded `.crb` would have been (see the [Disk size across fixture sizes](#disk-size-across-fixture-sizes) section below for hard numbers). The bpcells auto-integer-pack only kicks in for non-negative integer values (`slot = "counts"`); for normalised float values (`slot = "data"` / `"scale.data"`) the exporter falls back to raw double storage to avoid silent truncation. **The h5 win on RAM and load time stays consistent across fixture sizes; the disk picture is more nuanced.**

# Data

All benchmark data is **synthetic** — generated reproducibly by `tests/smoke/src/smoke_fixture_utils.R` and written to `tests/smoke/data/` by `tests/smoke/src/00_prepare_synthetic_data.R`. Five fixtures are generated, spanning a range of sizes typical for scRNA-seq experiments:

| Fixture | Genes | Cells | Density | nnz | Description |
|---|---|---|---|---|---|
| Myeloid | 4,000 | 8,000 | 2.5% | 0.8 M | snRNA-seq myeloid subset |
| Fibro Ctrl | 2,500 | 6,000 | 3.0% | 0.5 M | Xenium spatial fibroblasts |
| Fibro MS | 2,500 | 6,000 | 3.0% | 0.5 M | Xenium spatial fibroblasts |
| PBMC VDJ | 5,000 | 15,000 | 0.8% | 0.6 M | PBMC with TCR/BCR repertoire |
| **PBMC All** | **38,606** | **147,756** | **2.0%** | **114 M** | Large integrated PBMC |

The PBMC All fixture (38,606 genes × 147,756 cells, 114M non-zero entries) is the primary benchmark target — large enough that the three backends pull apart on every axis we measure. Smaller fixtures compress the differences too aggressively to be informative. The counts are drawn from Poisson(λ=3-5)+1, mimicking the small-count distribution of scRNA-seq data.

# What the benchmark measures

For each backend the script measures:

| metric | what it captures |
|---|---|
| `disk_crb_mb` | size of the `.crb` itself |
| `disk_sibling_mb` | size of the external sibling (NA for `embedded`) |
| `load_secs` | `readRDS()` of the `.crb` file — pure deserialisation of the metadata/projections/etc. plus, for `embedded`, the expression matrix itself |
| `attach_secs` | `.attachExternalExpression()` — for `embedded` it just validates the backend tag and returns; for `bpcells` it opens a lazy `IterableMatrix` handle on the sibling directory; for `h5` it opens a lazy `HDF5Array::TENxMatrix` seed and applies a free `DelayedArray::t()` to expose Cerebro's genes × cells layout (no data is read at attach time) |
| `rss_mb` | process resident-set size after load + attach |
| `cold_secs` | first single-gene full-row read after attach |
| `hot_p50_secs` / `hot_p95_secs` | 30 single-gene reads rotating through a 50-gene pool |
| `bulk_secs` | densified 50-gene × all-cells slice (the `marker_genes` tab pattern) |

Each backend runs in a **fresh R subprocess** via `callr::r()`, so the RSS reading for one backend is not polluted by the matrix left over from the previous backend. Reads go through the class accessors `getExpressionRow()` / `getExpressionBlock()` — the same code path the Shiny server uses — so the comparison reflects realistic runtime cost rather than raw `dgCMatrix` indexing.

# Running it locally

All scripts live under `tests/smoke/src/` and read fixtures from `tests/smoke/data/` (paths relative to `tests/smoke/`). First generate the synthetic data, then convert to all three backends, then run the benchmarks:

```{r}
# from tests/smoke/
Rscript src/00_prepare_synthetic_data.R      # generate synthetic fixtures
Rscript src/10_convert_embedded.R            # embedded backend
Rscript src/11_convert_bpcells.R             # bpcells backend
Rscript src/12_convert_h5.R                  # h5 backend
Rscript src/93_bench_backend_compare.R       # server-side bench
Rscript src/96_bench_size_scaling.R          # controlled size-scaling experiment
```

The bench produces `result/93_bench_backend_compare/summary.csv` (server side) and `result/96_bench_size_scaling/summary.csv` (size scaling). All fixture dimensions can be scaled via the `SMOKE_SYNTH_SCALE` environment variable (default 1.0).

# Results — PBMC All (38,606 × 147,756, 114M nnz)

Numbers below were captured on the PBMC All synthetic fixture, each backend in its own fresh R subprocess via `callr::r()`. Smaller is better in every column.

| metric | embedded | bpcells | h5 |
|---|---:|---:|---:|
| disk_crb_mb | 353.6 | 6.7 | **6.3** |
| disk_sibling_mb | — | 211.5 | **202.0** |
| **disk_total_mb** | 353.6 | 218.2 | **208.3** |
| load_secs | 3.293 | 0.389 | **0.368** |
| attach_secs | **0.014** | 0.538 | 0.432 |
| **startup_total_secs** | 3.307 | 0.927 | **0.800** |
| **rss_mb** | 2,289 | 1,035 | **1,015** |
| cold_secs | 0.2070 | 0.3163 | **0.0105** |
| hot_p50_secs | 0.1855 | 0.2933 | **0.0102** |
| hot_p95_secs | 0.2061 | 0.3384 | **0.0179** |
| bulk_secs (50 × ncells) | **0.245** | 0.740 | 0.511 |

# Disk size across fixture sizes

The h5-vs-embedded disk ranking depends on fixture size. The table below shows all five synthetic fixtures — note the crossover:

| Fixture | Dims | embedded | bpcells | h5 | h5 vs emb |
|---|---|---|---|---|---|
| Myeloid | 4k × 8k | **3.7 MB** | 8.4 MB | 6.9 MB | **+86%** |
| Fibro Ctrl | 2.5k × 6k | **3.9 MB** | 6.4 MB | 5.6 MB | **+44%** |
| Fibro MS | 2.5k × 6k | **3.9 MB** | 6.4 MB | 5.6 MB | **+44%** |
| PBMC VDJ | 5k × 15k | 2.5 MB | 2.6 MB | **2.4 MB** | −4% |
| **PBMC All** | 38.6k × 148k | 353.6 MB | 218.2 MB | **208.3 MB** | **−41%** |

On the three smallest fixtures (≤ 4k genes, ≤ 15k cells) the HDF5 chunk + dimnames + indptr overhead exceeds gzip's compression gain — h5 ends up 44-86% _larger_ than embedded. The crossover happens around 5k genes × 15k cells (PBMC VDJ), and by the time you reach a full-scale integration (PBMC All, 38.6k × 148k, 114M nnz) h5 is 41% _smaller_ than embedded.

This pattern is controlled and reproducible — the script `tests/smoke/src/96_bench_size_scaling.R` sweeps across three synthetic sizes (500 × 2k, 50k × 20k, optionally 500k × 20k) with identical sparsity and value distributions, isolating the backend disk scaling from any dataset-specific confounders:

| Size | Backend | total_mb | export_secs |
|---|---|---|---|
| small (500 × 2k) | embedded | **0.35** | 0.12 |
| small (500 × 2k) | bpcells | 0.31 | 0.25 |
| small (500 × 2k) | h5 | 0.31 | 0.16 |
| medium (50k × 20k) | embedded | 138.0 | 33.3 |
| medium (50k × 20k) | bpcells | **79.5** | 1.4 |
| medium (50k × 20k) | h5 | 83.8 | 20.7 |

At small scale all three are effectively tied (≤ 0.35 MB). At medium scale (50k cells × 20k genes, 50M nnz) the external backends pull ahead — bpcells and h5 are ~1.7× smaller than embedded on disk.

# Interpretation

**Disk** — `h5` wins on total disk _on this fixture_ (208 MB) because HDF5's default gzip filter compresses the TENx CSC arrays well at this scale; `bpcells` is close behind (218 MB) since 1.7.0 auto-converts integer counts to `uint32_t` before write, triggering BPCells's bit-packed integer storage (~5× smaller than raw double). `embedded` is the largest (354 MB) — saveRDS-serialised dgCMatrix. The `.crb` itself is tiny (~6-7 MB) for both external backends because it carries only metadata, projections, and trees plus a backend pointer — no expression matrix. The h5 disk win is fixture-size dependent: on small or dense fixtures HDF5 metadata + chunk overhead can flip the ranking and the h5 sibling ends up larger than the embedded crb (see the [Disk size across fixture sizes](#disk-size-across-fixture-sizes) section).

**Startup** — the total here is `load_secs + attach_secs`:

- `embedded` spends almost all of its budget in `load_secs` (~3.3 s) because the matrix is _inside_ the `.crb` and `readRDS()` has to deserialise the whole thing. `attach` is effectively free (0.014 s) — there is nothing external to attach, the helper just confirms the backend tag.
- `bpcells` has a small `load_secs` (~0.39 s, the `.crb` is only metadata) and a small `attach_secs` (~0.54 s). `open_matrix_dir()` only opens file handles on the sibling directory; no expression data is read until a query needs it.
- `h5` has the same small `load_secs` as `bpcells`, and `attach_secs` is also small (~0.43 s) because `HDF5Array::TENxMatrix(file, group)` only reads the top-level dimensions and dimnames; the actual `data`/`indices`/`indptr` arrays stay on disk and are read lazily at query time.

Net effect: `h5` and `bpcells` are essentially tied on startup (~0.8-0.9 s), both ~4× faster than `embedded` (~3.3 s).

**Memory** — `h5` and `bpcells` both run in ~1.0-1.1 GB RSS because neither materialises the full matrix; `h5` queries stream individual columns through `HDF5Array`'s page cache, `bpcells` queries stream chunks through its bit-packed integer format. `embedded` keeps the full sparse matrix in memory (~2.3 GB on this fixture).

**Query latency** — `h5` is dramatically faster than the other two on single-gene reads (~0.01 s, vs ~0.19 s for embedded and ~0.29 s for bpcells). Two effects compound here: (1) TENx CSC stores cells as rows and genes as columns, so a single-gene read is a contiguous CSC column slice, the cheapest possible HDF5 access pattern, and (2) the kernel page cache holds the relevant column blocks in RAM after the first read, so "hot" reads are basically memory-speed without paying the dgCMatrix indexing overhead embedded incurs. `bpcells` is slowest on single-gene reads because its row-oriented packed format has to densify chunks; the gap closes but doesn't disappear on bulk slices (50 genes × all cells).

# When to pick what

- **`h5`** _(recommended default for most deployments)_ — fastest startup, lowest RAM, fastest per-gene queries (~18× faster than embedded on hot reads). On disk: h5 is **smaller than embedded at typical scale** (≥ tens of thousands of cells × genes, where gzip compression wins) but _larger_ than embedded on very small or low-gene fixtures (HDF5 chunk + dimnames overhead dominates). The TENx CSC layout aligns with how Cerebro reads expression (per-gene → per-column), and HDF5 page-caching makes repeated reads memory-fast without committing the whole matrix to RAM. Cost: requires the `HDF5Array` Bioconductor package on the host. Export is slower (~21s on 50k×20k) due to single-threaded gzip, but this is a one-time cost at conversion time.
- **`bpcells`** — RAM-constrained host, very large matrices, or when you want BPCells's chunk-level optimisations (e.g. for batched all-cell operations rather than per-gene reads). Fastest export (~1.4s on 50k×20k). Per-gene query is ~0.3s, ~29× slower than h5 for this access pattern.
- **`embedded`** — single-file convenience (no sibling to manage), or compatibility with very old `.crb` readers. Best for small datasets or one-shot scripts that load and discard. Fastest bulk reads (~0.25s for 50 genes × all cells) since the `dgCMatrix` is already in RAM.

# Reproducing on your own data

`tests/smoke/src/93_bench_backend_compare.R` reads the PBMC All `.crb` files produced by scripts 10/11/12. To bench a different dataset, copy the script, swap the three `crb` / `sibling` paths in the `backends` list at the top, and rerun. The script gracefully skips backends whose `.crb` is missing, so partial setups (e.g. only embedded and h5) work too.

```{r}
backends <- list(
  embedded = list(
    crb     = "result/10_convert_embedded/cerebro_my_dataset.crb",
    sibling = NULL
  ),
  bpcells = list(
    crb     = "result/11_convert_bpcells/cerebro_my_dataset.crb",
    sibling = "result/11_convert_bpcells/cerebro_my_dataset.bpcells"
  ),
  h5 = list(
    crb     = "result/12_convert_h5/cerebro_my_dataset.crb",
    sibling = "result/12_convert_h5/cerebro_my_dataset.h5"
  )
)
```

# Reference

- `tests/smoke/src/smoke_fixture_utils.R` — synthetic data generator (`smoke_fixture_profile()`, `random_sparse_counts()`)
- `tests/smoke/src/00_prepare_synthetic_data.R` — writes synthetic fixtures to `tests/smoke/data/`
- `tests/smoke/src/{10,11,12}_convert_*.R` — converters per backend
- `tests/smoke/src/93_bench_backend_compare.R` — server-side bench (callr-isolated, writes `summary.csv`)
- `tests/smoke/src/96_bench_size_scaling.R` — controlled size-scaling experiment
- `?exportFromSeurat`, `?convertSeuratToCerebro` — backend selection via `expression_matrix_mode`

# Appendix: 1.6.0 eager h5 baseline

Before the 1.7.0 refactor the h5 attach was eager: `.attachExternalExpression()` called `rhdf5::h5read()` on all six datasets under `/expression/` and rebuilt a full `dgCMatrix` in RAM, mirroring the eager dgCMatrix path that `embedded` already uses. The disk format itself was a self-rolled CSC layout written via `rhdf5::h5write()`, not TENx — but the field names and on-disk shape happened to match TENx closely enough that `HDF5Array::TENxMatrix()` could read them after the fact.

The 1.7.0 refactor changed two files:

- `R/exportFromSeurat.R` — exporter now writes via `HDF5Array::writeTENxMatrix()` (canonical TENx schema, future-proof).
- `inst/shiny/v1.4/utility_functions.R` — attach now opens a lazy `HDF5Array::TENxMatrix` seed and lazily `t()`s it back to genes × cells (no data read at attach time; the in-memory `dgCMatrix` is never materialised).

The `Cerebro_v1.3` accessors `getExpressionRow()` / `getExpressionBlock()` already had a `DelayedMatrix` code path, so no downstream changes were needed in the class or the Shiny modules.

Same fixture, same scripts, same hardware — only the h5 attach implementation changed:

| metric | h5 (1.6.0 eager) | h5 (1.7.0 lazy) | improvement |
|---|---:|---:|---:|
| `attach_secs` | 22.9 | **0.087** | ~263× faster |
| `rss_mb` after attach | 11,210 | **1,133** | ~10× smaller |
| `cold_secs` (single gene) | 0.45 | **0.010** | ~45× faster |
| `hot_p50_secs` | 0.46 | **0.009** | ~51× faster |
| `bulk_secs` (50 × ncells) | 0.51 | **0.157** | ~3× faster |
| open URL → dataset visible | 33.0 s | **8.7 s** | ~4× faster |
| `disk_total_mb` | 354 | 391 | +10% (TENx adds a small metadata overhead) |

The `.crb` size (~42 MB) and `load_secs` (~3.3 s) are unchanged — only the attach path and everything downstream of it improved.

The 1.6.0 row above came from running `93_bench_backend_compare.R` against commit `b8cdadc`. To reproduce, `git checkout b8cdadc -- R/exportFromSeurat.R inst/shiny/v1.4/utility_functions.R`, regenerate the h5 sibling (with `12_convert_h5.R`), and rerun the bench.
